---
output: github_document
title: "unsurv: Unsupervised Clustering of Individualized Survival Curves"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# unsurv

<!-- badges: start -->

<!-- CRAN status (activate after submission) -->

<!-- [![CRAN status](https://www.r-pkg.org/badges/version/unsurv)](https://CRAN.R-project.org/package=unsurv) -->

<!-- R CMD check -->

<!-- [![R-CMD-check](https://github.com/USERNAME/unsurv/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/USERNAME/unsurv/actions) -->

<!-- License -->

[![License: MIT](https://img.shields.io/badge/license-MIT-blue.svg)](LICENSE)

<!-- badges: end -->

`unsurv` provides tools for *unsupervised clustering of individualized survival curves* using a medoid-based (PAM) algorithm. It is designed for applications where each observation is represented by a full survival probability trajectory over time, such as:

* survival model predictions
* individualized survival curves from Cox or deep learning models
* dynamic risk trajectories
* multi-omics prognostic profiles

The package provides:

* Medoid-based clustering of survival curves
* Automatic selection of number of clusters via silhouette
* Support for weighted L1 and L2 distances
* Prediction of cluster membership for new curves
* Stability assessment via resampling and Adjusted Rand Index
* Visualization tools


# Installation

## From GitHub

```{r, eval=FALSE}
install.packages("remotes")
remotes::install_github("ielbadisy/unsurv")
```

## From CRAN (after submission)

```{r, eval=FALSE}
install.packages("unsurv")
```


# Overview

The core function is:

```{r, eval=FALSE}
unsurv()
```

which clusters survival curves represented as an `n × m` matrix:

* rows = individuals
* columns = survival probabilities at time points


# Example: Clustering survival curves

```{r example-fit}
library(unsurv)

set.seed(123)

n <- 100
Q <- 50
times <- seq(0, 5, length.out = Q)

rates <- c(0.2, 0.5, 0.9)
group <- sample(1:3, n, TRUE)

S <- sapply(times, function(t)
  exp(-rates[group] * t)
)

S <- S + matrix(rnorm(n * Q, 0, 0.01), nrow = n)
S[S < 0] <- 0
S[S > 1] <- 1

fit <- unsurv(S, times, K = NULL, K_max = 6)

fit
```


# Plot cluster medoids

```{r example-plot}
plot(fit)
```

Each line represents the medoid survival curve for a cluster.

# Predict cluster membership for new curves

```{r example-predict}
predict(fit, S[1:5, ])
```


# Stability assessment

Cluster stability can be evaluated using resampling:

```{r example-stability}
stab <- unsurv_stability(
  S, times, fit,
  B = 20,
  frac = 0.7,
  mode = "subsample"
)

stab$mean
```

Higher values indicate more stable clustering.


# Using ggplot visualization


```{r example-autoplot, eval=requireNamespace("ggplot2", quietly=TRUE)}
library(ggplot2)
library(unsurv)
autoplot(fit)
```


# Methodological details

Given survival curves:

$$
S_i(t_1), S_i(t_2), \dots, S_i(t_m)
$$

the algorithm:

1. Optionally enforces monotonicity
2. Applies weighted feature transformation
3. Computes pairwise distances (L1 or L2)
4. Applies PAM clustering
5. Selects optimal K via silhouette (if not specified)

Cluster medoids represent prototype survival profiles.

# Typical workflow

```{r, eval=FALSE}
fit <- unsurv(S, times)

clusters <- fit$clusters

pred <- predict(fit, new_S)

stab <- unsurv_stability(S, times, fit)
```


# Applications

**unsurv** is useful for:

* Patient stratification
* Risk phenotype discovery
* Prognostic subgroup identification
* Deep survival model interpretation
* Precision medicine

# Relationship to other methods

Unlike clustering on covariates, **unsurv clusters on the survival function itself**, enabling:

* Direct interpretation of risk trajectories
* Robust clustering independent of feature space
* Model-agnostic analysis


# Package structure

Core functions:

| Function         | Description          |
| ---------------- | -------------------- |
| unsurv           | fit clustering model |
| predict          | assign new curves    |
| plot             | visualize medoids    |
| summary          | summarize clustering |
| unsurv_stability | evaluate stability   |
| autoplot         | ggplot visualization |


# Citation

If you use unsurv, please cite:

```
El Badisy, I. (2026). unsurv: Unsupervised clustering of individualized survival curves.
```


# License

MIT License © Imad EL BADISY


# Development status

[X] Stable core implementation
[X] Full test suite
[] CRAN submission ready


# Links

GitHub: [https://github.com/USERNAME/unsurv](https://github.com/USERNAME/unsurv)

